Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-04-21 Paper_Draft2/Draft2.Rmd
no more pkgs.bib

Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a sample COMPAS Risk Assessment Survey has been made publicly available, revealing the algorithm’s input information. This survey has been widely critiqued for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias. For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang. It also asks if a defendant has any friends or family members who have been crime victims [@Angwin2016Sample]. Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race. 
@angwin2016machine. published a 2016 piece in the news publication ProPublica that analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers the racial biases in defendants’ scores [@angwin2016machine]. Their research finds that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives. @angwin2016machine specifically investigate the distribution of COMPAS scores by decile among Black and white defendants. They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, Black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we]. After examining the fairness metric statistical parity difference, @angwin2016machine conclude that the algorithm is racially biased [@larson2016we].

@@

[Women at the Table](https://www.womenatthetable.net/), the sponsor organization for this project, is “a growing, global gender equality & democracy CSO based in Geneva, Switzerland focused on advancing feminist systems change by using the prism of technology, innovation & AI exercising leverage points in technology, the economy, sustainability & democratic governance.” We are collaborating with the organization on its AI & Equality [@noauthor_ai_nodate] initiative, tasked with de-biasing the COMPAS algorithm [@aif360-oct-2018] and producing a corresponding data story that will be added to its library. 
Our project builds on Women at the Table’s various de-biasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS data set. Based on this analysis, we employ a human rights framework to contribute to the ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm. With a solid understanding of the two sides, we aim to pinpoint the shortcomings of both arguments and correct them in our analyses. We will use various de-biasing techniques and fairness metrics to evaluate the level of bias present in the COMPAS data and our algorithm. We will summarize our results using the JupyterNotebook framework from Women at the Table, to be used by members of the organization to teach in a workshop setting. We hope that our findings will highlight the importance of checking statistical analyses using varied methods and contribute to the ongoing discussion of the effects of machine biases in the justice system.

>>>>

Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a [sample COMPAS Risk Assessment Survey](https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE#document/p5/a296598) has been made publicly available, revealing the algorithm’s input information. This survey has been widely critiqued for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias. For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang. It also asks if a defendant has any friends or family members who have been crime victims [@Angwin2016Sample]. Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race. 
@angwin2016machine. published a 2016 piece in the news publication ProPublica that analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers the racial biases in defendants’ scores [@angwin2016machine]. Their research finds that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives. @angwin2016machine specifically investigate the distribution of COMPAS scores by decile among Black and white defendants. They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, Black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we]. After examining the fairness metric statistical parity difference, @angwin2016machine conclude that the algorithm is racially biased [@larson2016we].

@@

[Women at the Table](https://www.womenatthetable.net/), the sponsor organization for this project, is “a growing, global gender equality & democracy CSO based in Geneva, Switzerland focused on advancing feminist systems change by using the prism of technology, innovation & AI exercising leverage points in technology, the economy, sustainability & democratic governance.” We are collaborating with the organization on its AI & Equality [@noauthor_ai_nodate] initiative, tasked with de-biasing the COMPAS algorithm [@aif360-oct-2018] and producing a corresponding data story that will be added to its library.

Our project builds on Women at the Table’s various de-biasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS data set. Based on this analysis, we employ a human rights framework to contribute to the ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm. With a solid understanding of the two sides, we aim to pinpoint the shortcomings of both arguments and correct them in our analyses. We will use various de-biasing techniques and fairness metrics to evaluate the level of bias present in the COMPAS data and our algorithm. We will summarize our results using the JupyterNotebook framework from Women at the Table, to be used by members of the organization to teach in a workshop setting. We hope that our findings will highlight the importance of checking statistical analyses using varied methods and contribute to the ongoing discussion of the effects of machine biases in the justice system.
